---
title: "age structure"
author: "ZHE ZHENG, Dan Weinberger, modified from Matlab code by Virginia Pitzer"
date: "12/26/2020"
output: html_document
---
```{r}
library(deSolve)
library(RColorBrewer)
library(reshape2)
```
In Pitzer et al: The <12 month olds, were divided into monthy age classes. The remaining population was divided into 6 classes: 1–4 years old, 5–9 years, 10–19 years, 20–39 years, 40–59-years, and 60+ years old. In this example, the 1-4 year olds are divided into 12 month age classes as well. 


```{r}
#T is N time points; N_ages is number of age groups
parmset <- readRDS('./data_and_parms/parmset.rds')
Pop1 <- readRDS('./data_and_parms/pop1.rds') #initial population
B <-  readRDS('./data_and_parms/Birth_rate.rds')
c2 <- readRDS( './data_and_parms/c2.rds') # contact matrix

```

#View contact matrix
```{r}
n.cols=100
nice.cols <-  colorRampPalette(brewer.pal(9, "YlOrRd"))(n.cols)
heatmap(c2/sum(diag(c2)), Rowv=NA, Colv=NA, scale='none', col=nice.cols)
```
Set names of Age groups 
```{r}
N_ages <- length(Pop1) 
agenames <- paste0('Agegrp', 1:N_ages) #Could replace this with vector of actual age names
```

```{r}
###########################################
# 1/duration of maternal immunity (DAYS)
DurationMatImmunityDays= 112
###########################################

###########################################
# update LRI and hospitalization probability
###########################################
al=21
delta1=c(rep(.40,3),rep(.39,3), rep(.21,3),rep(.20,3),0.16,rep(.14,3),rep(0.05,al-16))#proportion of first infections that are LRI (by age)
delta2=.5*delta1#proportion of second infectious that are LRI 
delta3=.7*delta2#proportion of third infections that are LRI
hosp1=c(.18*rep(.40,3),0.08*rep(.39,3),0.07*rep(.21,3),0.06*rep(.20,3),0.06*0.16,0.05*rep(.14,3),0.02*rep(0.05,al-16))#proportion of first LRIs that are hospitalized
hosp2=.4*hosp1#proportion of second LRIs that are hospitalized
hosp3=c(rep(0,al-2),0.00001,0.00004)#proportion of third LRIs that are hospitalized

```

## Population size by time
```{r}
p <- sum(Pop1)  # Total population at each time, a vector of length T
```

## Initialize the compartments (States) 
```{r}
StateNames <- c('M','S0','I1','S1','I2','S2','I3','S3','I4')
States <- array(NA, dim=c(N_ages, length(StateNames) )) #  N age groups xN parameters 
dimnames(States)[[1]] <- agenames
dimnames(States)[[2]] <- StateNames

yinit.matrix <- array(NA, dim=c(N_ages, length(StateNames) ))

dimnames(yinit.matrix)[[1]] <- agenames
dimnames(yinit.matrix)[[2]] <- StateNames

yinit.matrix[,c('S1','I2','S2','I3','S3','I4')]  = 0
yinit.matrix[,'M'] = c(Pop1[1:6], rep(0,N_ages-6))
yinit.matrix[,'S0'] = c(rep(0,6),Pop1[7:N_ages]-rep(N_ages-6)) 
yinit.matrix[,'I1'] = c(rep(0,6), rep(1,N_ages-6))  #initializes with 1 infected person per age group 

yinit.vector <- as.vector(yinit.matrix) #Vectorize the ynit matrix

# Create array that has the labels by age, State and use this to name the yinit.vector
name.array <- array(NA, dim=dim(yinit.matrix))
for(i in 1:dim(name.array)[1]){
  for(j in 1:dim(name.array)[2]){
      name.array[i,j] <- paste(dimnames(yinit.matrix)[[1]][i],dimnames(yinit.matrix)[[2]][j]  )
    }
  }

name.vector <- as.vector(name.array)
names(yinit.vector) <- name.vector

```


## Read in model and parameters

```{r}
source('./data_and_parms/simple_model.R') #Read in the model
```


**NOTE: time step here is in months--need to adjust seasonality accordingly**
```{r}
 start_time = 1 # start date (years)
 tmax = nrow(B)
# end_time = 25 # end date (years)
my_times <- seq(start_time, tmax, by = 1) # gives a sequence from start to end
											 # in increments of 1
```


##Settings parameters
Most of these are described in Table 2 of Pitzer et al, PLOS Pathogens: https://journals.plos.org/plospathogens/article?id=10.1371/journal.ppat.1004591

For birth rate:
The weekly per capita birth rate (B_w) is equal to log(1+B)/52.18 because the data on the birth rate is annual and the differential equation model inherently assumes that growth is occurring exponentially. So if the annual birth rate is equal to 12 per 1000 per year (B=0.012), for example, then we are assuming that:

$N1=N0exp(B_w*52.10)$     where N1 is the population after 1 year and N0 is the baseline population, and $N1=N0*(1+B)$.

$1.012=1*exp(B_w*52.18)$

$B_w=log(1.012)/52.18$.

```{r setparms}
#########################################
#Matrix of dimension N_ages x N_ages
#beta = beta 
#########################################

#########################################
#Seasonal components--
#???Should prob be estimated?
b1=0.1830827 #Seasonal amplitude
phi=3.327749 #Seasonal phase shift (weeks;[-26,26] 0=peak @ Jan 1)
#########################################

#########################################
#Relative infectiousness for 2nd and subsequent infections
rho1 = 0.75
rho2 = 0.51
#########################################

#########################################
# duration of infectiousness (months)
#Duration in days
dur.days1 <- 10 #days
dur.days2 <- 7 #days
dur.days3 <- 5 #days
###########################################  

###########################################
# 1/duration of maternal immunity (DAYS)
DurationMatImmunityDays= 16*7
###########################################

###########################################
WidthAgeClassMonth = c(rep(1,times=12), rep(12,times=4),  60, 120, 240, 240, 240 )  #Aging rate=1/width age class (months) Vector of long N_age
############################################

############################################
#????
#What is time(-1) unit?
#um= -0.0001833333  #?? is this right? only die from last age class--this maybe helps with this "We assumed deaths occurred from the last age class and adjusted the net rate of immigration/emigration and death from other age groups in order to produce a rate of population growth and age structure similar to that of the US." 
#?????????
#From Giny's code: um=log(0.993)/52 #net rate of crude deaths (+) and immigration (-) from all age groups (per week): can adjust this to approximate population growth in state# I calibrated this parameter so we can reproduce the population growth
um= -0.0002227
#############################################

#############################################
#Birth rate (births/person/YEAR
#Matrix: T rows, N_ages columns; columns 2:N_ages all 0s
PerCapitaBirthsYear=B 
#############################################

#############################################
#Relaive risk of infection following 1st, 2nd, 3rd+ infections
sigma1=0.76
sigma2=0.6
sigma3=0.4
#############################################


##To ESTIMTE from DATA?: Baseline transmission rate
#############################################

baseline.txn.rate <- 9.0
q=1
c2=c2

```


#Save parameters in a list
```{r saveparms}
parms<-list(PerCapitaBirthsYear=PerCapitaBirthsYear,
            DurationMatImmunityDays=DurationMatImmunityDays,
            WidthAgeClassMonth=WidthAgeClassMonth,
            um=um,
            b1=b1,
            phi=phi,
            rho1=rho1,
            rho2=rho2,
            dur.days1=dur.days1,
            dur.days2=dur.days2,
            dur.days3=dur.days3,
            yinit.matrix=yinit.matrix,
            baseline.txn.rate = baseline.txn.rate,
            q=q,
            c=c2,
            sigma1=sigma1,
            sigma2=sigma2,
            sigma3=sigma3,
            time.step='month'
            )
```


## Run the model 
Note here we are just simulating with set parameters, not fitting to data
```{r}
results <- ode(y=yinit.vector, t=my_times,  
               func=simple_model, 
            parms=parms )

```

## Extract the relevant results
```{r}

#Pitzer used a 40-50 YEAR burn in period
burnN <- 20*12
results.burned <- results[-c(1:burnN),]

##Any infected person
infected.cols <- results.burned[,c(grep('I1', colnames(results.burned)),grep('I2', colnames(results.burned)),grep('I3', colnames(results.burned)),grep('I4', colnames(results.burned)  )) ]

infected.all <- apply(infected.cols,1, sum)

infected.cols.m <- melt(infected.cols)
infected.cols.m$agegrp <- sub(" .*", "",infected.cols.m$Var2 )
infected.cols.c <- dcast(infected.cols.m, Var1~agegrp, fun.aggregate = sum)

#Pop
all.m <- melt(results.burned[,grep('Agegrp',colnames(results.burned))])
all.m$agegrp <- sub(" .*", "",all.m$Var2 )
all.c <- dcast(all.m, Var1~agegrp, fun.aggregate = sum)
pop.all <- apply(all.c[,-1],1,sum)

plot(infected.all, type='l')
plot(pop.all,type="l")

```
```{r}
  St <- results[-c(1:300),-1]
   q=1
   b= parms$baseline.txn.rate/(parms$dur.days1/30.44)
   beta=(b/100)/(sum(yinit.matrix)^(1-q))*c2
   b1= parms$b1
   phi=parms$phi
   rho1=parms$rho1
   rho2=parms$rho2
   sigma1=parms$sigma1
   sigma2=parms$sigma2
   sigma3=parms$sigma3
   t0=108
   I1 <- St[,grep('I1', colnames(St))]
   I2 <- St[,grep('I2', colnames(St))]
   I3 <- St[,grep('I3', colnames(St))]
   I4 <- St[,grep('I4', colnames(St))]
   S1 <- St[,grep('S1', colnames(St))]
   S2 <- St[,grep('S2', colnames(St))]
   S3 <- St[,grep('S3', colnames(St))]
   S0 <- St[,grep('S0', colnames(St))]
   lambda1=matrix(0,nrow=t0,ncol=al)#Force of infection
    for (t in 1:t0)
    {lambda1[t,] <- as.vector((1+b1*cos(2*pi*(t-phi*12)/12))*((I1[t,]+rho1*I2[t,]+rho2*I3[t,]+rho2*I4[t,])%*%beta)/sum(St[t,]))}

   
   H1=matrix(0,nrow=t0,ncol=al)#Number of hospitalizations by age
   for (i in 1:al){
     H1[,i]=hosp1[i]*S0[,i]*lambda1[,i]+hosp2[i]*sigma1*S1[,i]*lambda1[,i]+hosp3[i]*sigma2*S2[,i]*lambda1[,i]+hosp3[i]*sigma3*S3[,i]*lambda1[,i]
   }
   
```

```{r}
plot(1:108,rowSums(H1),type="l",xlab="months",ylab="simulated RSV hospitalizations")
# need to get rid of the burn-in period

barplot(c(sum(colSums(H1)[1:3]),sum(colSums(H1)[4:6]),
            sum(colSums(H1)[7:9]),
            sum(colSums(H1)[10:12]),
            colSums(H1)[13:21])/sum(H1),xlab="age group",ylab="Number of Hospitalizations")

# simulated age distribution Age groups: The <12 month olds, were divided into per 3-months age classes. The 1-4 year olds are  divided into 1yearly age classes, then 5–9 years, 10–19 years, 20–39 years, 40–59-years, and 60+ years old

Hosp_sim <- as.integer(rowSums(H1))
agedist_Sim <- c(sum(colSums(H1)[1:3]),sum(colSums(H1)[4:6]),
            sum(colSums(H1)[7:9]),
            sum(colSums(H1)[10:12]),
            colSums(H1)[13:21])/sum(H1)
```


## fit the model to simulated data in order to estimate parameter sets
```{r}
parmset<-list(PerCapitaBirthsYear=PerCapitaBirthsYear,
            WidthAgeClassMonth=WidthAgeClassMonth,
            um=um,
            rho1=rho1,
            rho2=rho2,
            dur.days1=dur.days1,
            dur.days2=dur.days2,
            dur.days3=dur.days3,
            yinit.matrix=yinit.matrix,
            q=q,
            c=c2,
            sigma1=sigma1,
            sigma2=sigma2,
            sigma3=sigma3,
            time.step='month'
            )

fitmodel <-  function(parameters,dat) {   # takes the parameter values and dataset as inputs
  protrans <- parameters[1] # estimate parameter related to R0 (baseline transmission rate)
  amp <- parameters[2] # estimate parameter related to seasonal amplitude
  trans <- parameters[3] # estimate parameter related to seasonal peak timing
  DMD <- parameters[4] # estimate parameter related to the duration of maternal immunity
  b1 <- exp(amp) #ensure positive 
  baseline.txn.rate <- exp(protrans) #ensure positive
  phi <-  (2*pi*(exp(trans))) / (1+exp(trans)) # transform to its scale in the model
  durx <- exp(DMD) #ensure positive
  durx <<- durx #somehow we need this for the function and model to recognize these parameters
  baseline.txn.rate <<- baseline.txn.rate
  b1 <<- b1
  phi <<- phi


   # Simulate the model with initial conditions and timesteps defined above, and parameter values from function call
  results <- ode(y=yinit.vector, t=my_times,  
               func=simple_model, 
            parms=c(parmset,baseline.txn.rate=baseline.txn.rate,b1,phi,DurationMatImmunityDays=durx))
  t0 <- 108
  results<- tail(results,t0)
  St <- results[,-1]
   
   b=baseline.txn.rate/(parmset$dur.days1/30.44)
   beta=(b/100)/(sum(yinit.matrix)^(1-1))*c2
 
   lambda1=matrix(0,nrow=t0,ncol=al)#Force of infection
    for (t in 1:t0)
    {lambda1[t,] <- as.vector((1+b1*cos(2*pi*(t-phi*12)/12))*((St[t,(2*al+1):(3*al)]+rho1*St[t,(4*al+1):(5*al)]+rho2*St[t,(6*al+1):(7*al)]+rho2*St[t,(8*al+1):(9*al)])%*%beta)/sum(St[t,]))}

  
  H1=matrix(0,nrow=t0,ncol=al)#Number of hospitalizations by age
  for (i in 1:al){
    H1[,i]=hosp1[i]*St[,(al+i)]*lambda1[,i]+hosp2[i]*sigma1*St[,(3*al+i)]*lambda1[,i]+hosp3[i]*sigma2*St[,(5*al+i)]*lambda1[,i]+hosp3[i]*sigma3*St[,(7*al+i)]*lambda1[,i]
    }
  H <- rowSums(H1)

  agedist <- c(sum(colSums(H1)[1:3]),sum(colSums(H1)[4:6]),
            sum(colSums(H1)[7:9]),
            sum(colSums(H1)[10:12]),
            colSums(H1)[13:21])/sum(H1)
   
   LLall <- sum(dpois(x = dat, lambda =H, log = TRUE)) # fit to timeseries
   LLmulti <- dmultinom(x= agedist,prob = agedist_Sim,log = T) # fit to age distribution
   
   #prior
   durprior <- dgamma(x=durx,22,5,log=T)

   #total LL
   LL <- LLall+LLmulti+durprior
   
   return(LL)
}

fitLL <- optim(par = c(2.198522,  -1.638387 , -1.301047 ,  4.619743 ),           # starting values for beta and gamma - you should get the same result no matter which values you choose here
      fn = fitmodel,        # the distance function to optimise
      dat = Hosp_sim,         # the dataset to fit to ("dat" argument is passed to the function specified in fn)
      control = list(fnscale=-1)) # the log likelihood is negative; here we minimize the negative log likelihood
```

